---
title: "TIDYVERSE PLOTLY"
author: "Jerald Melukkaran"
date: "2025-04-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(plotly))
```

#### Introduction 

In this vignette we observe how plotly can be used to animate plots which can be a great way to visualize changes in a dataset. Here i use it to visualize changes in genres of the top selling games across the years 2000 to 2015. You can either click play to automatically run through the years, or use the slider. 


```{r warning=FALSE}

library(plotly)
library(dplyr)


vgsales <- read.csv("https://raw.githubusercontent.com/jerryjerald27/Data-607/refs/heads/main/Tidyverse/vgsales.csv")

vgsales_filtered <- vgsales %>% filter(Year >= 2000 & Year <= 2015)

plot_ly(data = vgsales_filtered, 
        x = ~Global_Sales, 
        y = ~Genre, 
        color = ~Genre, 
        frame = ~Year, 
        type = 'bar', 
        orientation = 'h', 
        marker = list(line = list(width = 10))) %>%
  layout(
    xaxis = list(title = 'Global Sales (millions)'),
    yaxis = list(title = 'Game Genre'),
    title = 'Top Selling Game Genres (2000 to Present)',
    updatemenus = list(
      list(
        type = 'buttons',
        showactive = FALSE,
        buttons = list(
          list(method = 'animate', args = list(NULL, list(frame = list(duration = 1000, redraw = TRUE), mode = 'immediate')))
        )
      )
    )
  )

```

Animating the plots can really help visualize the changes. We will use another example here. We will animate the graphs to show the changes in the sales of each platform (Xbox, Playstation, Nintendo, etc) over the same 15 year span.

```{r}

vgsales_by_platform <- vgsales_filtered %>%
  group_by(Platform, Year) %>%
  summarise(Global_Sales = sum(Global_Sales), .groups = 'drop')


plot_ly(data = vgsales_by_platform, 
        x = ~Global_Sales, 
        y = ~Platform, 
        color = ~Platform, 
        frame = ~Year, 
        type = 'bar', 
        orientation = 'h',  # Horizontal bars
        marker = list(line = list(width = 2))) %>%
  layout(
    xaxis = list(title = 'Global Sales (millions)'),
    yaxis = list(title = 'Platform'),
    title = 'Global Sales of Each Platform (2000 to 2015)',
    showlegend = FALSE,  # Hide the legend
    updatemenus = list(
      list(
        type = 'buttons',
        showactive = FALSE,
        buttons = list(
          list(method = 'animate', args = list(NULL, list(frame = list(duration = 1000, redraw = TRUE), mode = 'immediate')))
        )
      )
    )
  )
```

